Clinical advancement forecasting \thanks{\textit{\underline{Citation}}: \textbf{TBD}}

### 2024-04-11 paper/paper.tex
Add pagebreaks in supplements
Finish initial Methods draft

\section{Introduction}

@@

The features used throughout this study consist of 27 target-disease pair predictors, 5 target-specific predictors and 1 disease-specific predictor. These are listed in Table~\ref{tab:features}. The target and disease specific features are chosen carefully such that they are either capable of being associated with years in which events supporting them occurred or result from large-scale, unbiased methods that do not favor well-studied or drugged targets/diseases. Examples of this include target-specific tissue expression specificity scores computed from Human Protein Atlas \cite{PMID:25613900} and LOEUF \cite{PMID:32461654} scores from gnomAD. Simply put, our dataset combines scores from Open Targets for target-disease evidence and a select subset of target prioritisation \cite{OT23.12release} fields with almost no modifications, other than to add target and disease specific indicators of maximum trial phases reached and two extra genetic association features (described in Section~\ref{sec:methods}).

\begin{figure}[!htb]
  \centering
  \captionsetup{width=.9\linewidth}
  \captionsetup[subfigure]{labelformat=empty}
  \subfloat[\centering]{
    \begin{overpic}[width=.9\textwidth]{feature_presence_unscaled.pdf}
      \put(102, 20) {(a)}
    \end{overpic}
  }
  \qquad
  \subfloat[\centering]{
    \begin{overpic}[width=.9\textwidth]{feature_presence_scaled.pdf}
      \put(102, 20) {(b)}
    \end{overpic}
  }
  \caption{
    \textbf{Evaluation dataset feature presence}.
    (a) Presence and co-occurrence of various feature groupings or individual features indicated by overlapping sets of TD pairs, where the size of each bar is proportional to the number of TD pairs with a given combination of features and the sort ordering is determined by the number of TD pairs associated with any one feature (show in the right margin). The top margin shows the number of overlapping features for a combination of features and the bottom margin shows the number of TD pairs that are a part of that same combination.
    (b) Same as (a) without truly proportional bar sizes that are not scaled down beyond a limit necessary to fit all margin labels and combinatorial groupings. Features that are not specific to TD pairs are also omitted for brevity, along with the outcome feature used in this study and denoted here with the label "pair has advanced beyond phase 2".
  }
  \label{fig:feature_presence}
\end{figure}

@@

The use of such a metric is essential for properly assessing performance in this forecasting problem. While we also report more common measures of classifier performance like Receiver Operating Characteristic (ROC) and Average Precision (PR), neither of these adequately capture behavior in the upper extremes of rankings due the sparsity with which TD pair evidence is present for pairs that have ever entered phase 2 trials. This sparsity is further exacerbated in this study by the constraint that most of that evidence must have existed \textit{before} such trials began. For more details on the extent of this sparsity, see Section~\ref{sec:features}. The figure presented there, Figure~\ref{fig:feature_presence}, also demonstrates that a substantial fraction (78\%) of TD pairs that ultimately advance beyond phase 2 trials have targets and/or diseases with prior clinical validation despite no direct evidence linking the pairs themselves, which means a comprehensive measure of classifier performance (e.g. ROC) is far more likely to reflect the extent to which disease-only historical, clinical information or target-only information -- including other attributes like conservation, essentiality and tissue expression -- can predict clinical success. Again, this is not our primary focus as we want to evaluate the maximum achievable performance in this forecasting problem, and we assert that this is best accomplished when one or more lines of target-disease-specific evidence are present.

@@

@@

One key objective of this study is to determine if any model, e.g. RDG, can sort TD pairs with genetic support such that at least some portion of that sorted list has a likelihood of advancement that consistently exceeding what is expected from any one source of genetic support alone. We find that this goal is met and exceeded by the RDG model, which actually identifies more TD pairs than those that have either EVA or GWAS support alone at an expected rate of advancement exceeding that of the single source (respectively). This does not appear to be the case with the OMIM baseline, however the lack of examples in our evaluation dataset with OMIM support makes any determination difficult. See Section~\ref{sec:results_opportunities} for more on how these benchmarks are employed to contextualize opportunities among undeveloped TD pairs and Supplementary~Figure~\ref{fig:top_evaluation_predictions} for top predictions from the RDG model. This latter, supplementary figure further emphasizes our focus on prioritizing opportunities beyond those with genetic support and provides examples of TD pairs with multiple lines of evidence.

\begin{figure}[!htb]
  \centering
  \captionsetup{width=.9\linewidth}
  \begin{overpic}[width=1\textwidth]{relative_risk_by_limit.pdf}
    % Ranking percentages
    \put(23,8){\line(0,1){2}}
    \put(20,11){Top 1\%}
    \put(37,8){\line(0,1){2}}
    \put(34,11){Top 2\%}
  \end{overpic}
  \caption{
    \textbf{Performance compared to genetic support benchmarks}. The RR estimates for each benchmark annotated on the right are shown as horizontal dotted lines and are calculated across all TD pairs (N=\evaluationDatasetPairCount). The number of pairs with support from each benchmark is represented by the bars extending horizontally (to the right) from the y-axis. The RDG RR point estimate is shown in blue and bounds around that estimate correspond to a Katz 90\% confidence interval. Annotations on the x-axis indicate how many TD pairs comprise the top 1\% (N=90) and top 2\% (N=180) of rankings. All results shown are from the evaluation dataset.
  }
  \label{fig:relative_risk_by_limit}
\end{figure}

We also note that Supplementary~Figure~\ref{fig:relative_risk_core_features} shows confidence intervals for each of the genetic benchmarks of Figure~\ref{fig:relative_risk_by_limit} in isolation, as well as all other target-disease-specific evidence sources, in addition to confidence intervals for the RDG model at various top ranking cutoffs. Similar comparisons for target-specific and disease-specific features can be seen in Supplementary~Figure~\ref{fig:relative_risk_static_features}. These findings suggest that 1) genetic support for TD pairs is highly predictive but rare, 2) human clinical support for targets and diseases in isolation is also predictive while being more common and 3) constraint and expression specificity of targets exhibit modest but significant effects. While only human clinical evidence appears to have a prognostic value rivaling that of genetic support when considered on a univariate basis, the combined influence of non-genetic, non-clinical evidence is examined in Section~\ref{sec:model_comparison} where a model using only this information, RDG-X, still outperforms Open Targets composite scores by all measures.

@@

\begin{figure}[!htb]
	\centering
  \captionsetup{width=.9\linewidth}
	\includegraphics[width=1\textwidth]{effect_sizes.pdf}
  \caption{
    \textbf{RDG model feature effects}. The \textbf{effect max} values are equivalent to RDG model coefficients for the corresponding feature while the \textbf{effect mean} values indicate average values of the product between the coefficient and a particular feature value, when that feature is present.
  }
	\label{fig:effect_sizes}
\end{figure}

@@

A common method for identifying druggable opportunities within a specific disease context involves first ranking TD pairs according to some prioritization methodology followed by filtering or reprioritizing those ranks based on knowledge of target tractability \cite{PMID:28356508,PMID:35401535,PMID:31253980}. We use a similar approach to identify tractable targets associated with TD pairs that have yet to enter clinical trials. To aid in interpreting this approach, we also draw on the results of Figure~\ref{fig:relative_risk_by_limit}. The data in this figure suggests thresholds for the RDG model that align to expected rates of advancement compared to several genetic support benchmarks. These thresholds are used to bucket undeveloped TD pairs before further bucketing them based on levels of tractability. The tractability buckets in Supplementary~Table~\ref{tab:tractability_buckets} provide \colorbox{Gainsboro}{HIGH}, \colorbox{Gainsboro}{MED}, and \colorbox{Gainsboro}{LOW} confidence ratings for each type of tractability evidence based on the priorities suggested in \cite{OTTractability}.

@@

@@

% TODO: complete this section?
Another way to quantify the impact that employing superior target prioritization strategies could have is to consider what would happen if this was done across the entire pharmaceutical development industry. Our internal research shows that \$130B (92\%) of global pharma R\&D costs (\$150B) are expended on failed assets each year. This assumes an 8\% success rate that if improved by 3x to 24\%, would lead to a total cost savings of approximately \$22B annually ...

Throughout the majority of this study, we focus primarily on performance among the top \topRankingPct\% of TD pairs. Reasons for this are introduced in Section~\ref{sec:features} and Section~\ref{sec:metrics}. It is constructive to expand on this choice by comparing it to alternatives, and noting that it is quite likely that many other predictors could improve performance for the remaining \bottomRankingPct\%. We offer some evidence of this through the use of a feature indicating how long a TD pair has been in phase 2 trials, which improves wholistic measures of ranking performance like ROC by 12 points (see Figure~\ref{fig:performance_metric_mean_across_ta}). While time is a necessary but not sufficient condition for clinical advancement of a TD pair, given that a single trial can take years to complete, it also correlates with how many distinct trials, drugs, companies, sponsors, investigators, etc. attempt to validate any one pair. We posit that this is crucial because even if a druggable, mechanistic link exists between a target and a disease, many (or even most) trials testing that link will fail for reasons unrelated to efficacy or safety of a drug \cite{Razuvayevskaya2023.02.07.23285407}. Some of these reasons include 1) commercial factors like a lack of funding, pipeline reprioritization and competitive density, 2) administrative or logistical factors like a lack of enrollment, retention problems, poor trial design, drug supply chain shortages, epidemics (COVID-19) and 3) regulatory factors like legislative shifts or extensive regulatory approval delays. Many of these factors are fundamentally difficult or impossible to predict. Epidemic outbreaks, labor shortages leading to supply chain problems and regulatory shifts are examples of exogenous shocks for which accurate predictors are unlikely to exist. However, factors like funding, competition, enrollment and certain aspects of trial design have measurable predictors such as preceding venture capital and private equity investments, patent filings and revenue for related drugs, disease severity and prevalence, and the existence of biomarkers or other enabling factors for better trial designs (respectively). As a more detailed example, patient recruitment failures are the most common reason trials fail \cite{Razuvayevskaya2023.02.07.23285407} and recruitment statistics are highly predictive of eventual trial outcomes \cite{Lo2019Machine,PMID:34430930}. It stands to reason then that predicting trial enrollment failures well in advance of when a trial is even conceived, i.e. for a target-disease pair, should be possible because recruitment in trials is determined, in part, by how prevalent a disease is, how debilitating it is, what patient demographics it inflicts, whether existing treatments exist for it and how successful recruitment for past trials (if any exist) has been -- all of which are measurable. Signals like this could be used to further improve performance in the bottom \bottomRankingPct\% of TD pairs since they better capture causal factors in non-biological trial failures that time under development -- the only non-biological feature in this study -- does not.

Another notable focus of this study is on evidence for TD pairs that can be \textbf{directly} attributed to them. This is a departure from earlier research in this space like \cite{Nelson2015-eg} that often includes measures of similarity between disease ontology terms to account for the fact that bridging disease nomenclatures used in clinical trial datasets with those used in databases maintaining genetic evidence is difficult. It is not uncommon for disease terms from either source to be sufficiently similar such that it is appropriate to consider them as equivalent. It can also be argued that similarity between disease terms offers an important dimension for expanding evidence in a biologically meaningful way, regardless of technical mapping issues. This is taken to further extremes in studies like \cite{PMID:33262371} that propagates evidence using both measures of target similarity (from protein interaction networks) and measures of disease similarity (from ontologies and literature co-occurrence). We believe such expansion methods are very compatible with our approach by either using them to derive new predictors or to propagate predictions from models like these across networks or knowledge graphs.

\begin{itemize}
  \item Discuss the possibility to do prospective evaluation with OT snapshots
  \item Cover limitations with OT concerning temporalization for both evidence and drug approvals, and why only transitions from phase 2 are relevant for this work
  \item Discuss the limitations of the tractability analysis and why they might not be tractable targets
  \item From \cite{PMID:33262371}: "It is important to bear in mind therefore that what we are measuring when looking at historical trial outcomes is not an unbiased measure of any given gene's true disease associations, but rather a view on how useful a given evidence source or analytical method has been for choosing drug targets based on current and historical drug discovery practices. Dramatic changes in these practices in the future could render some of our conclusions obsolete, though the fundamental observation that genetic association itself is retained in molecular networks will remain valid."
\end{itemize}

@@

\begin{itemize}
  \item Discuss why all target prioritisation data fields are not used due to the potential leakage they may impose (e.g. target families, GO annotations, etc.)
  \item Describe how therapeutic areas were selected based on having at least 100 TD pairs with target-disease-specific evidence of any kind and with explicit omissions: ("biological\_process", "pregnancy or perinatal disease", "injury, poisoning or other complication", "pregnancy or perinatal disease", "medical procedure", "infectious disease", "animal disease")
  \item RDG models are fit using the \colorbox{Gainsboro}{Ridge} implementation from scikit-learn and GBM models are fit using the LightGBM \cite{LightGBM} algorithm
  \item Cite supervenn
  \item Training/evaluation results are limited by temporalization while the present-day predictions are not
  \item OMIM is defined as EVA associations with publications
  \item The "genetic\_association\_\_curated" field is a union of all genetic association sources other than "gene\_burden" and "ot\_genetics\_portal"
  \item Mention that the scores for a TD pairs in a year are the maximum score for that source, not the harmonic sum
  \item Mean imputation is used for target-specific features
  \item Mention specifics on RDG and GBM implementations, i.e. lightGBM and scikit-learn
  \item RDG is always trained on all features, but only applied to feature subsets where relevant
\end{itemize}

@@

\subsection{Features}

@@

\begin{figure}[!htb]
  \centering
  \captionsetup{width=.9\linewidth}
  \includegraphics[width=1\textwidth]{dataset_statistics.png}
  \caption{
    \textbf{Training and evaluation dataset summary statistics}.
  }
  \label{fig:dataset_statistics}
\end{figure}

@@

\subsection{Performance}

@@

@@

\subsection{Predictions}

@@

\subsection{Opportunities}

@@

@@

\subsection{Sensitivity}

>>>>

\section{Introduction}
\label{sec:introduction}

@@

The features used throughout this study consist of 27 target-disease pair predictors, 5 target-specific predictors and 1 disease-specific predictor. These are listed in Table~\ref{tab:features}. The target and disease specific features are chosen carefully such that they are either capable of being associated with years in which events supporting them occurred or result from large-scale, unbiased methods. This choice was motivated in part by empirical results presented in Section~\ref{sec:results_inflation} and assumptions described more in Section~\ref{sec:methods}. Simply put, our dataset combines scores from Open Targets for target-disease evidence and a select subset of target prioritisation \cite{OTtargetPrioritisation} fields with almost no modifications, other than to add target and disease specific indicators of maximum trial phases reached and two extra genetic association features (also described in Section~\ref{sec:methods}).

\begin{figure}[!htb]
  \centering
  \captionsetup{width=.9\linewidth}
  \captionsetup[subfigure]{labelformat=empty}
  \subfloat[\centering]{
    \begin{overpic}[width=.9\textwidth]{feature_presence_unscaled.pdf}
      \put(102, 20) {(a)}
    \end{overpic}
  }
  \qquad
  \subfloat[\centering]{
    \begin{overpic}[width=.9\textwidth]{feature_presence_scaled.pdf}
      \put(102, 20) {(b)}
    \end{overpic}
  }
  \caption{
    \textbf{Evaluation dataset feature presence}.
    (a) Presence and co-occurrence of select features for overlapping sets of TD pairs. The size of each bar is proportional to the number of TD pairs with a given combination of features and the sort ordering is determined by the number of TD pairs associated with any one feature (show in the right margin). The top margin shows the number of features in any one feature combination and the bottom margin shows the number of TD pairs that are a part of that same combination.
    (b) Same as (a) with proportional bar scaling that does not drop beyond a limit necessary to fit all margin labels and combinatorial groupings. Features that are not specific to TD pairs are omitted for brevity, along with the outcome feature used in this study and denoted here with the label "pair has advanced beyond phase 2".
  }
  \label{fig:feature_presence}
\end{figure}

@@

The use of such a metric is essential for properly assessing performance in this forecasting problem. While we also report more common measures of classifier performance like Receiver Operating Characteristic (ROC) and Average Precision (PR), neither of these adequately capture behavior in the upper extremes of rankings due the sparsity with which TD pair evidence is present for pairs that have ever entered phase 2 trials. This sparsity is further exacerbated in this study by the constraint that most of that evidence must have existed \textit{before} such trials began. For more details on the extent of this sparsity, see Section~\ref{sec:features}. The figure presented there, Figure~\ref{fig:feature_presence}, also demonstrates that a substantial fraction (78\%) of TD pairs that ultimately advance beyond phase 2 trials have targets and/or diseases with prior clinical validation despite no direct evidence linking the pairs themselves, which means a comprehensive measure of classifier performance (e.g. ROC) is far more likely to reflect the extent to which disease-only historical, clinical information or target-only information -- including other attributes like conservation and tissue expression -- can predict clinical success. Again, this is not our primary focus as we want to evaluate the maximum achievable performance in this forecasting problem, and we assert that this is best accomplished when one or more lines of target-disease-specific evidence are present.

@@

We conclude that the RDG model outperforms OTS by a factor of at most 2 and least 1.5 for a significant number of TD pairs. A sensitivity analysis for this key result is presented in Supplementary~Section~\ref{sec:sensitivity}, where we find that these multipliers hold on average across multiple Open Targets releases and other important settings.

@@

One key objective of this study is to determine if any model, e.g. RDG, can sort TD pairs with genetic support such that at least some portion of that sorted list has a likelihood of advancement that consistently exceeds what is expected from any one source of genetic support. We find that this goal is met and surpassed by the RDG model, which actually identifies more TD pairs than those that have either EVA or GWAS support and have an equivalent or greater expected rate of advancement. This does not appear to be the case with the OMIM baseline, however the lack of examples in our evaluation dataset with OMIM support makes any determination difficult. See Section~\ref{sec:results_opportunities} for more on how these benchmarks are employed to contextualize opportunities among undeveloped TD pairs and Supplementary~Figure~\ref{fig:top_evaluation_predictions} for top predictions from the RDG model. This latter, supplementary figure further emphasizes our focus on prioritizing opportunities beyond those with genetic support and provides examples of TD pairs with multiple lines of evidence.

\begin{figure}[!htb]
  \centering
  \captionsetup{width=.9\linewidth}
  \begin{overpic}[width=1\textwidth]{relative_risk_by_limit.pdf}
    % Ranking percentages
    \put(23,8){\line(0,1){2}}
    \put(20,11){Top 1\%}
    \put(37,8){\line(0,1){2}}
    \put(34,11){Top 2\%}
  \end{overpic}
  \caption{
    \textbf{Performance compared to genetic support benchmarks}. The RR estimates for each benchmark annotated on the right are shown as horizontal dotted lines and are calculated across all TD pairs (N=\evaluationDatasetPairCount). The number of pairs with support from each benchmark is represented by the bars extending horizontally (left to right) from the y-axis. RDG RR point estimates are shown in blue and bounds around the estimates correspond to Katz 90\% confidence intervals. Annotations on the x-axis indicate how many TD pairs comprise the top 1\% (N=90) and top 2\% (N=180) of rankings. All results shown are from the evaluation dataset.
  }
  \label{fig:relative_risk_by_limit}
\end{figure}

We also note that Supplementary~Figure~\ref{fig:relative_risk_core_features} shows confidence intervals for each of the genetic benchmarks of Figure~\ref{fig:relative_risk_by_limit} in isolation, as well as all other target-disease-specific evidence sources, in addition to confidence intervals for the RDG model at various top ranking cutoffs. Similar comparisons for target-specific and disease-specific features can be seen in Supplementary~Figure~\ref{fig:relative_risk_static_features}.

@@

\begin{figure}[!htb]
	\centering
  \captionsetup{width=.9\linewidth}
  \begin{overpic}[width=1\textwidth]{effect_sizes.pdf}
    \put(48, .5) {(a)}
    \put(70, .5) {(b)}
  \end{overpic}
  \caption{
    \textbf{RDG model feature effects}. 
    (a) Feature average effects calculated as the mean of the product between a coefficient and a particular feature value, when that feature is present and non-negative (i.e. when it has any influence). Only the evaluation dataset is used to compute this average.
    (b) Feature coefficients, which are equivalent to the maximum effect that a feature can have due to all features being defined on the range [0, 1] or rescaled to it.
  }
	\label{fig:effect_sizes}
\end{figure}

@@

A common method for identifying druggable opportunities within a specific disease context involves first ranking TD pairs according to some prioritization methodology followed by filtering or re-prioritizing those ranks based on knowledge of target tractability \cite{PMID:28356508,PMID:35401535,PMID:31253980}. We use a similar approach to identify tractable targets associated with TD pairs that have yet to enter clinical trials. To aid in interpreting this approach, we also draw on the results of Figure~\ref{fig:relative_risk_by_limit}. The data in this figure suggests thresholds for the RDG model that align to expected rates of advancement compared to several genetic support benchmarks. These thresholds are used to bucket undeveloped TD pairs before further bucketing them based on levels of tractability. The tractability buckets in Supplementary~Table~\ref{tab:tractability_buckets} provide \colorbox{Gainsboro}{HIGH}, \colorbox{Gainsboro}{MED}, and \colorbox{Gainsboro}{LOW} confidence ratings for each type of tractability evidence based on the priorities suggested in \cite{OTTractability}.

@@

We also note that these TD pairs are identified using predictors that are not constrained by temporalization. In other words, the predictions used to identify \textbf{present-day} TD pairs with a high likelihood of clinical success are based on more evidence than what is available in other sections of this study. This is primarily because some evidence in Open Targets is not associated with publications. Ignoring that limitation is appropriate in this context.

@@

% TODO: complete this section?
Another way to quantify the impact that employing superior target prioritization strategies could have is to consider what would happen if this was done across the entire pharmaceutical development industry. Our internal research shows that \$130B (92\%) of global pharma R\&D costs (\$150B) are expended on failed assets each year. This assumes an 8\% success rate that if improved by 3x to 24\%, would lead to a total cost savings of approximately \$22B annually ... (WIP)

Throughout the majority of this study, we focus primarily on performance among the top \topRankingPct\% of TD pairs. Reasons for this are introduced in Section~\ref{sec:features} and Section~\ref{sec:metrics}. It is constructive to expand on this choice by comparing it to alternatives, and noting that it is quite likely that many other predictors could improve performance more comprehensively. We offer some evidence of this through the use of a feature indicating how long a TD pair has been in phase 2 trials, which improves wholistic measures of ranking performance like ROC by 12 points (see Figure~\ref{fig:performance_metric_mean_across_ta}) and provides a particularly dramatic lift across the bottom \bottomRankingPct\% of TD pairs. While time is a necessary but not sufficient condition for clinical advancement of a TD pair, given that a single trial can take years to complete, it also correlates with how many distinct trials, drugs, companies, sponsors, investigators, etc. attempt to validate any one pair. We posit that this is crucial because even if a druggable, mechanistic link exists between a target and a disease, many (or even most) trials testing that link will fail (or not even start) for reasons unrelated to efficacy or safety of a drug \cite{Razuvayevskaya2023.02.07.23285407}. Some of these reasons include 1) commercial factors like a lack of funding, pipeline reprioritization and competitive density, 2) administrative or logistical factors like a lack of enrollment, retention problems, poor trial design, drug supply chain shortages, epidemics (COVID-19) and 3) regulatory factors like legislative shifts or extensive regulatory approval delays. Many of these factors are fundamentally difficult or impossible to predict. Epidemic outbreaks, labor shortages leading to supply chain problems and regulatory shifts are examples of exogenous shocks for which accurate predictors are unlikely to exist. However, factors like funding, competition, enrollment and certain aspects of trial design have measurable predictors such as preceding venture capital and private equity investments, patent filings and revenue for related drugs, disease severity and prevalence, and the existence of biomarkers or other enabling factors for better trial designs (respectively). As a more detailed example, patient recruitment failures are the most common reason trials fail \cite{Razuvayevskaya2023.02.07.23285407} and recruitment statistics are highly predictive of eventual trial outcomes \cite{Lo2019Machine,PMID:34430930}. It stands to reason then that predicting trial enrollment failures well in advance of when a trial is even conceived, i.e. for a target-disease pair, should be possible because recruitment in trials is determined, in part, by how prevalent a disease is, how debilitating it is, what patient demographics it inflicts, whether existing treatments exist for it and how successful recruitment for past trials has been -- all of which are measurable. Signals like this could be used to further improve performance since they better capture causal factors in non-biological advancement failures that time under development -- the only non-biological feature in this study -- does not.

Another notable focus of this study is on evidence for TD pairs that can be \textbf{directly} attributed to them. This is a departure from earlier research in this space like \cite{Nelson2015-eg} that often includes measures of similarity between disease ontology terms to account for the fact that bridging disease nomenclatures used in clinical trial datasets with those used in databases maintaining genetic evidence is difficult. It is not uncommon for disease terms from either source to be sufficiently similar such that it is appropriate to consider them as equivalent. It can also be argued that similarity between disease terms offers an important dimension for expanding evidence in a biologically meaningful way, regardless of technical mapping issues. This is taken to further extremes in studies like \cite{PMID:33262371} that propagates evidence using both measures of target similarity (e.g. using protein interaction networks) and measures of disease similarity (e.g. using ontologies and literature co-occurrence). We believe such expansion methods are very compatible with our approach by either using them to derive new predictors or to propagate predictions from models like these across networks or knowledge graphs.

All results we have presented are based on quasi-prospective outcomes. As described in Section~\ref{sec:introduction}, this means that the outcomes are not truly observed in the future relative to the information used to predict them. Both outcomes and predictors are assigned timestamps corresponding to publication or database dates that approximately capture when the events associated with them actually occurred. The accuracy of that approximation is difficult to determine, and publication dates can be particularly inaccurate when they lag evidence discovery by months or years. This does not pose a risk for information leaks in a quasi-prospective formulation though, as long as the date associated with an event does not precede its occurrence. Nevertheless, it is possible that publication dates are simply entered into a database incorrectly, clinical trial registries inaccurately represent when trials began or ended, or that information from future events ends up associated with the wrong targets and/or diseases. These are only a few possible risks for information leaks, and though each is relatively rare on its own, their combined impact could be substantial. And as with all forecasting models, the accuracy of future predictions is reliant on the extent to which the underlying data generating process may either change or deviate from anticipated changes. Fortunately, the evidence we employ as predictors here comes predominantly from molecular interactions in humans, animals or cells that are likely to remain stable in time, or shift on evolutionary time scales that eclipse drug development timelines. This summary on the subject from \cite{PMID:33262371} is pertinent:

\begin{quotation}
  It is important to bear in mind therefore that what we are measuring when looking at historical trial outcomes is not an unbiased measure of any given gene's true disease associations, but rather a view on how useful a given evidence source or analytical method has been for choosing drug targets based on current and historical drug discovery practices. Dramatic changes in these practices in the future could render some of our conclusions obsolete, though the fundamental observation that genetic association itself is retained in molecular networks will remain valid.
\end{quotation}

Model performance evaluations based on prospective outcomes, as an alternative to quasi-prospective outcomes, offer a number of distinct advantages. For one, they mitigate nearly all the risks for information leaks mentioned previously by ensuring that all predictive information precedes clinical outcomes. They also offer a means to quantify those risks. An extension of this work might utilize the Open Targets database snapshots, dating back as far as 2016, to examine how much model performance differs between models trained on features from a snapshot and models trained on data from a current Open Targets release with features that are temporalized and then filtered to not exceed the year of the earlier snapshot. Similarly, that extension could examine how much model performance differs for a snapshot with and without longitudinal features by considering only outcomes that occurred after the snapshot. Such an approach could also be applied on a per-feature basis to gain more granular insight on which kinds of evidence pose the greatest risk. One clear disadvantage of prospective evaluation for this problem is that it requires years to accrue enough future clinical outcomes to be meaningful. This also means that predictors must omit that many years of accrued evidence as well, which can deflate performance compared to what is possible in the present day. Another disadvantage is that it is significantly more difficult to operationalize prospective evaluations, thereby implying that the decrease in risk of information leaks may be partially offset by an increase in technical risk.

@@

All data used in this study comes from Open Targets with one minor exception. Open Targets provides PubMed identifiers as provenance for evidence without supplying publication dates or years in all cases. These dates are taken from \cite{PMID:33763309} instead. Our primary results are based on the 23.12 Open Targets release, but we also use releases 23.09 and 23.06 as part of our sensitivity analysis in section \ref{sec:sensitivity}.

The features we derive from Open Targets come primarily from a single target-disease dataset containing only "direct" evidence \cite{OTdirectVsIndirect}. These features are named according to the \colorbox{Gainsboro}{datasourceId} field within this dataset, and the values for those features are defined as the cumulative, maximum score observed when ordering those scores by the years associated with the evidence underlying them. We also add 2 genetic features that are not immediately available. The first, \colorbox{Gainsboro}{genetic\_association\_\_curated} is a union of all genetic association sources other than \colorbox{Gainsboro}{gene\_burden} and \colorbox{Gainsboro}{ot\_genetics\_portal}. The second, \colorbox{Gainsboro}{genetic\_association\_\_omim}, is defined as any EVA associations with a publication date.

The target-specific features we derive come from the "target prioritisation" dataset \cite{OTtargetPrioritisation}. The only exception to this is a feature defining the maximum trial phase reached, which comes from the primary target-disease evidence dataset. Most scores in this dataset were not used because they cannot be temporalized. We make exceptions for those arising from large-scale, genome-wide assays like LOEUF from gnomAD \cite{PMID:32461654} and tissue expression scores from Human Protein Atlas \cite{PMID:25613900}. We retain a score derived from phenotype severity observed in mice knock-outs \cite{PMID:33231642} as well. All three of these feature sources are unique in that they are imputed differently from all other features when absent in model training. The "closed-world" assumption discussed in Section~\ref{sec:results_inflation} implies that a constant, zero imputation for most missing features is appropriate. This is not true for non-clinical, target-specific fields -- they are mean-imputed instead.

All ridge regression models (e.g. RDG) are fit using the scikit-learn \colorbox{Gainsboro}{Ridge} implementation and all gradient-boosted machine models (e.g. GBM-T) are fit using the LightGBM \cite{LightGBM} algorithm. Ridge models are fit with very weak regularization (\colorbox{Gainsboro}{alpha=1e-6}) and LightGBM models are fit with default settings. Variants of both models are fit, in some cases, with non-negative monotonicity constraints. An important caveat unique to our use of the ridge regression models is that when report results for feature ablations, as in Section~\ref{sec:model_comparison}, all features are actually used to train the models before using only the unablated features to make predictions.

Most results reported in this study consider only a subset of therapeutic areas within the Experimental Factor Ontology (EFO) \cite{PMID:20200009}. All therapeutic areas are retained if they possessed at least 100 TD pairs with target-disease-specific evidence of any kind within our evaluation dataset. We also explicitly omit the "biological\_process", "pregnancy or perinatal disease", "injury, poisoning or other complication", "pregnancy or perinatal disease", "medical procedure", "infectious disease", and "animal disease" therapeutic areas, as they are either not relevant or poorly represented. The "infectious disease" therapeutic area is a borderline case that we omit not because it is necessarily irrelevant, but because it has only 115 associated TD pairs. This makes it the most poorly represented therapeutic area that is possibly relevant. We remove it by explicit omission and use a 100 TD pair threshold rather than using a 115 TD pair threshold.

\section{Limitations}

There are a number of limitations and caveats with this study. These include:

\begin{enumerate}
  \item Open Targets does not currently provide drug approvals with associated dates. This is a part of the reason why we have chosen to focus only on advancement beyond phase 2, and not phase 3 as well. 
  \item Not all Open Targets evidence can be associated with publication dates. This evidence is ignored when used to construct longitudinal predictors; however, this limitation does not apply to Section~\ref{sec:results_opportunities}.
  \item The Open Targets composite score we use (OTS) is not equivalent to the overall composite score for TD pairs provided by Open Targets \cite{OTscoring}. The score they provide is not longitudinal. We create an approximation of it over time based on the same weighting scheme across evidence types, but without a harmonic sum aggregation function. Maximum scores for the same target, disease, year and evidence type combination are used instead.
  \item Estimating an absolute number of TD pairs associated with any criteria is difficult. In Section~\ref{sec:results_opportunities}, we present absolute counts of undeveloped TD pairs with tractable targets and a high likelihood of clinical success. These counts may be inflated by the fact that the associated targets and diseases are so sufficiently similar that they ought not to be counted as distinct examples. Defining a threshold for similarity to combat this problem is difficult and application-specific though, so we do not offer any solutions here.
  \item The source of OMIM associations we use is derived from a subset of associations available in EVA. This subset is defined by associations with a linked publication. This is only an approximation, but our internal research shows that 98\% of EVA records with linked publications also have at least one OMIM submission.
\end{enumerate}

\pagebreak

@@

\begin{figure}[H]
  \centering
  \captionsetup{width=.9\linewidth}
  \includegraphics[width=1\textwidth]{dataset_statistics.png}
  \caption{
    \textbf{Training and evaluation dataset summary statistics}.
  }
  \label{fig:dataset_statistics}
\end{figure}

\pagebreak

@@

\clearpage

@@

\pagebreak

@@

\pagebreak

@@

\pagebreak

@@

\pagebreak

@@

\clearpage

@@

\pagebreak

\subsection{Sensitivity}
\label{sec:sensitivity}
